Inferring causal connectivity from pairwise recordings and optogenetics

### 2018-09-16 manuscript/main.tex
removed all deprecated code and results and renamed all notebooks to respective figure names

\documentclass[11pt]{article}
\usepackage{amsmath,array}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{breakcites}
\usepackage{graphicx,subcaption}
\usepackage{todonotes} %[disable]
\usepackage{cleveref}
\usepackage{tikz}
\usepackage{placeins} %FloatBarrier
\usepackage{booktabs}
\usepackage{natbib}
\usetikzlibrary{arrows,shapes,calc}
\tikzset{line/.style = {draw, -latex', ultra thick}}
\newcommand{\abs}[1]{\left|{#1}\right|}
\newcommand{\pde}[2]{{\frac{\partial {#1}}{\partial {#2}}}}
\newcommand{\pdde}[2]{\frac{\partial^2 {#1}}{\partial {#2}^2}}
\newcommand{\de}[2]{\frac{\mathrm{d} #1}{\mathrm{d} #2}}
\newcommand{\dde}[2]{\frac{\mathrm{d}^2 {#1}}{\mathrm{d} {#2}^2}}
\newcommand{\intdef}[4]{\int\limits_{#1}^{#2} #3\, \mathrm{d}{#4}}
\newcommand{\mean}[1]{\left<{#1}\right>}
\crefname{appendix}{Appendix}{appendices}
\crefname{section}{Section}{sections}
\crefname{table}{Table}{tables}
\Crefname{table}{Table}{Tables}
\crefname{equation}{Eq.}{eqs.}
\Crefname{equation}{Equation}{Equations}
\crefname{figure}{Fig.}{figs.}
\Crefname{figure}{Figure}{Figures}
\crefname{subfigure}{Fig.}{figs.}
\Crefname{subfigure}{Figure}{figures}
\captionsetup[subfigure]{labelformat=simple}
\renewcommand{\thesubfigure}{(\alph{subfigure})}
\renewcommand{\vec}[1]{\boldsymbol{#1}}
%opening
\title{Inferring causal connectivity from pairwise recordings and optogenetics}
%\title{The need to combine optogenetics with neural recordings to estimate causal connectivity}
\author{Mikkel Elle Lepper\o{}d, Konrad Paul Kording}

@@

\begin{abstract}
\noindent Neuroscientists often want to identify how neurons causally interact. Perturbations with optogenetics are a popular way of quantifying such interactions. However, even focused light will typically stimulate multiple cells, which produces a confound - we cannot know which of the stimulated neurons affected the activity of a given target neuron. Here we show how this confound produces large biases in estimates of connectivity, and how we can reduce these biases by a data analysis strategy that exploits the absolute refractory period of neurons. We use the instrumental variable (IV) framework from economics which estimates causal strengths by exploiting a variable that affects only one of the variables, here the presynaptic neuron. The interaction between optogenetic stimulation and the absolute refractory period produces a weak almost random signal (IV) that we exploit to solve the causal inference problem. In a simple simulation of integrate and fire neurons, we find that instrumental variable techniques estimate connectivity better than naive techniques (99.8\% vs 86\% correct). Our instrumental variable approach is robust and easy to implement in the context of current experimental paradigms.
\end{abstract}

\section{Introduction}
% neuroscience experiments are observational which implicates a certain set of statistical tools
% spike timing cannot give a causal estimate of connectivity, thus we need to perturbe the system
%a quest for causality and optogenetics is a main tool
% We want mechanism. Seriously. Neuroscience is to a large part experimentally driven.
%set up the drive to causality
We want to understand the mechanisms or causal chains that give rise to activity in the brain, to perception, action, and cognition. For such an understanding it is not sufficient to know the correlations between variables or even be able to predict them. After all, there can be many ways how the same activities come about by distinct causal chains \citep{drton2011global, peters2017elements}. Identifiability of causal networks is complicated when it comes to brain data. Complex systems such as the brain are hard to understand because of the numerous ways the contributing elements may interact internally \citep{Jonas2017}. Thus, brain data which is almost always observational, will almost never satisfy the criteria needed for identifiability \citep{pearl2009causality}. While observing correlations within the system is usually relatively easy, transitioning from observed correlations to a causal or mechanistic understanding is hard or, more commonly, impossible.  Getting at such an understanding in the human brain is nearly impossible as it contains approximately 86 billion neurons \citep{azevedo2009equal}, each of which influences many other neurons. Only under certain assumptions about nonlinearity or noise sources does a fully observed system become identifiable \citep{daniusis2012inferring,shimizu2006linear}. Even if we could record all neurons at the same time, estimating causality and producing a mechanistic understanding would be extremely challenging.

@@

%optogenetics is totally nonlocal
Interpreting the results of optogenetic stimulation in terms of causal interactions is difficult. The expression of opsins are often restricted to a population of neurons, but regular optogenetic stimulation will still affect many neurons simultaneously. In \textit{in vivo} settings it is generally impossible to direct photons to only one neuron. Hence, the stimulus will produce a distributed pattern of activity. For example, if we focus a stimulation beam on one neuron, there will be a cone of light in front of and behind the neuron, which comes from the rays of light from the lens focused on the cell and again coming out on the backside.. This distributed pattern of stimulation produces activity which then percolates through the network of neurons. Thus any post-synaptic activity induced by stimulation could in principle come from any of the stimulated neurons introducing problematic confounders.

@@

\section{Results}
\subsection{Optogenetics is not local}
Optogenetics is generally seen as a perturbation method that by and large affects neurons in close proximity of the light source. However, it can be questioned if this is the correct way of conceptualizing the spatial effect of stimulation. The stimulation effects depend on multiple factors. Firstly, light intensity and opsin density are important as more light and ion channels will cause a stronger effect on each cell. Secondly, the number of potentially stimulated neurons is critical as more neurons will have a larger impact on the overall population activity. Lastly, physiological properties of the cells are important as light may e.g. only have a strong effect on spiking activity when the membrane potential of the cell is sufficiently close to the firing threshold. The induced effect of optogenetic stimulation as a function of distance should be the product of four parameters:  light intensity, spatial distribution of neurons, distributions of membrane potential across neurons, and opsin distribution across neurons.

To estimate the light intensity we calculated the spatial extent of laser light delivered by fiber-optics under plausible experimental conditions according to \cite{Aravanis2007}; see \cref{sec:method:opto}. This modeling of light intensity yield an approximately $ 1/r^2 $ reduction with distance $ r $ from the stimulation site \cref{fig:concept} (cyan line). This is explained by the surface of the 3D shell growing with $4\pi r^2$ and photons will be roughly isotropic beyond the scattering length \cref{fig:concept} (inset). The same number of photons has to cross each of the spheres around the stimulation location unless they are absorbed. The density of photons thus decreases rapidly with distance.

@@

To estimate the effect of stimulation, the functionality underlying spiking activity needs to be considered. This can largely be characterized by the  distribution of membrane potentials across neurons. Surprisingly, this distribution has been observed to be symmetrically distributed and relatively flat \citep{pare1998impact,destexhe1999impact,rudolph2006use}. The response that is expected due to a pulse of light that induces a charge $Q$ should be proportional to the density of neurons whose membrane potential sit within a $Q/C$ range of the threshold ($C$ is the capacitance). The fact that the distribution of membrane potentials is relatively flat (the density close to the threshold is generally within an order of magnitude of the density of its mode) suggests that the response in membrane potential to a perturbation for any neuron is roughly proportional to the light intensity. Moreover, assuming that opsins are evenly distributed along the neuron, the opsin density in the neuron population will be relatively even. Based on this, we calculate the overall stimulation effect to be the product of neuron density and light intensity which is approximately constant in distance (up to the distance where absorption becomes important) \cref{fig:concept} (blue line). Thus, optogenetic stimulation utilizing single photon activation does not actually produce a localized effect.

\begin{figure}\includegraphics[scale=1]{opto-powerlaw}
\caption{{\bf Spatial extent of optogenetic stimulus}. Due to scattering and geometric loss the light intensity (I, cyan line) plotted as percentage of intensity exiting the optogenetic fiber follows approximately an inverse square law $ r^{-2} $ where $ r $ is the distance from the fiber. If neurons are uniformly distributed the number of affected neurons increase by $ r^{2} $ (N, black line) rendering the probability of activating a neuron approximately constant (IN, blue line). \label{fig:concept}}
\end{figure}

\subsection{Confounding as a problem for the estimation of causal effects}
When we stimulate many neurons at the same time, and observe a post-synaptic neuron to be active after our stimulation, it is hard to know which of the stimulated neurons produced the activity. To illustrate such confounding effects we simulated a network comprised of three neurons (A,B, and C). The neurons receive Poisson spike trains and have added Gaussian white noise to the membrane potential. We also gave them interactions, where spikes of neuron B increase the probability of firing for neuron C but there are no other interactions \cref{fig:intro}\labelcref{fig:intro:1}.
Finally, we allowed simulated optogenetic stimulation to affect neurons A and B (but not C). We thus have a simple system for exploring questions of causality.

@@

\medskip
\begin{subfigure}{\textwidth}\centering \includegraphics[scale=1]{xcorr}
\caption{} \label{fig:intro:3}
\end{subfigure}
\caption{{\bf A,B,C network}. A sketch of the simple network containing three neurons %depicted in \labelcref{fig:intro:3} 
shows stimulation configuration with blue laser light and the connections with arrows (a). The neurons A and B are stimulated 1000 trials and the corresponding peristimulus time histogram are shown in \labelcref{fig:intro:2} upper panel with a raster plot in the lower panel. Cross correlation histograms (CCHs) are shown in \labelcref{fig:intro:3} where horizontal and vertical axes represents time lag in ms and counts of coincident spikes in bins of 1 ms. \label{fig:intro}}
\end{figure}

\subsection{Instrumental variables to resolve confounding}
In order to estimate the actual influence of stimulation of a neuron on post-synaptic neurons we need something that can distinguish the influence of one stimulated neuron from the influence of another stimulated neuron. We would thus need something that affects the stimulation effect separately between neurons. Arguably, refractoriness is such a variable. If a neuron is in its absolute refractory period, then no amount of stimulation will make it spike. This gives us an interesting way of getting at causality, by comparing the network state between a time when a neuron is able to spike and a time where the neuron is unable to spike.

@@

We can now investigate if the use of an instrumental variable gives a better estimate of connectivity strength than simply analyzing the lagged correlations by means of the CCH; see \cref{eq:ptrans}. We thus use the IV estimator \cref{eq:wald} on the three neuron system (\cref{fig:cchvswald}\labelcref{fig:cchvswald:1,fig:cchvswald:2}). It converges to the correct causal conclusions that the weights $ w_{BC} = 0.2 $ and $ w_{AC} = 0 $. For such a simple system, it produces meaningful estimates of the causal interactions between neurons.

\begin{figure}
\makeatletter
\renewcommand\p@subfigure{}
\makeatother
\begin{subfigure}{0.485\textwidth} \centering
% * <tristan.stoeber@posteo.net> 2018-06-24T21:23:23.987Z:
% What is S?
% ^.
\begin{tikzpicture}
    \node (u) {$ S $};
    \node [above = of u] (x) {$ A/B $};
    \node [above right = of u] (y) {$ C $};
    \node [left = of x] (sr) {$ A_r/B_r $};
    \path [line] (x) -- (y);
    \path [line] (u) -- (x);
    \path [line] (u) -- (y);
    \path [line] (sr) -- (x);
\end{tikzpicture}
\caption{} \label{fig:cchvswald:0}
\end{subfigure}\hfill
\begin{subfigure}{0.485\textwidth} \centering
\includegraphics[scale=1]{wald_legend}
\end{subfigure}\medskip

\begin{subfigure}{0.485\textwidth} 
\includegraphics[scale=1]{wald_BC}
\caption{} \label{fig:cchvswald:1}
\end{subfigure}\hfill
\begin{subfigure}{0.485\textwidth} 
\includegraphics[scale=1]{wald_AC}
\caption{} \label{fig:cchvswald:2}
\end{subfigure}
\caption{{\bf Instrumental variable estimation (IV) of connectivity}. \labelcref{fig:cchvswald:0} In an instrumental variable estimation procedure we use a variable that is assumed to be random (here refractoriness) which influences a variable of interest (here spiking) and to use this influence to infer the causal interaction of that variable on other variables (here spiking of A or B onto C). \labelcref{fig:cchvswald:1} A popular estimation approach for IVs,  the Wald technique correctly estimates causal connectivity in the A,B,C neural network using the refractory period. Subfigure \labelcref{fig:cchvswald:0} shows a path diagram \citep{wright1921correlation} between the upstream neuron $ A $ or $ B $, the stimulation $ S $, the post-synaptic neuron $ C $ and the IV as $ A_{r} $ or $ B_{r} $. Arrows represent associations where $ S $ is associated with $ A,B $ and potentially also with $ C $ both directly and through $ A,B $. The IV estimator calculated by \cref{eq:wald} converges to $ \hat{\beta}_{BC} \approx 0.2, \hat{\beta}_{AC} \approx 0 $ after approximately $ 5000 $ trials as seen in \labelcref{fig:cchvswald:1,fig:cchvswald:2} respectively. % I would have put in a take home message / interpretation of the results in the figure text.
Insets represent high resolution zoom of cross correlation histograms of BC and AC where horizontal and vertical axes represents time lag and counts of coincident spikes in bins of 0.1 ms respectively. \label{fig:cchvswald}}
\end{figure}

\subsection{Larger simulated networks}
The interacting neurons in a biological network exhibit inhibition and interact in many kinds of ways. To evaluate the IV method in a more meaningful setting we thus simulated a recurrent neural network consisting of 1250 randomly connected leaky integrate and fire neurons where 250 had inhibitory synapses. The network was tuned to be in an asynchronous regime; see \cref{fig:state}\labelcref{fig:CC} with log-normally distributed synaptic weights according to patch clamp experiments \citep{Sayer1990,Mason1991}; see \cref{fig:state}\labelcref{fig:syn_dist} and \cref{tab:params} for parameters. Further, we selected 800 excitatory neurons for stimulation and gave each neuron a random spatial distance from the simulated optogenetic stimulus. The stimulus intensity was then set according to \cref{eq:intensity} with a maximum of 10 pA, and was constant throughout trials. The trial onset had a temporal Poisson distribution with period 100 ms and was further clipped between 100-150 ms. For weight estimates we randomly selected among the excitatory population 50 stimulated neurons and 50 unstimulated neurons.

@@

\begin{subfigure}{\textwidth} 
\includegraphics[scale=1]{mse_pos}
\caption{} \label{fig:error:mse}
\end{subfigure}
\caption{{\bf Mean square error (MSE) of IV estimator in a large network}. \labelcref{fig:error:raster} Excitatory neurons (in blue raster) are stimulated with varying intensity indicated by the y-axis in lower panel, upper panel shows inhibitory neurons (red raster) where y-axis indicate random neuron-id. \labelcref{fig:error:mse} The IV- and CCH estimator is evaluated for the asynchronous recurrent neural network at three different amounts of relative inhibition $ g $. The MSE as a function of number of trials is shown on a logarithmic scale where the slopes was found to be $ -0.77, -0.48, -0.40 $ for $g = 9.9, 4.4, 3.0$ respectively. \label{fig:error}}
\end{figure}

We want to compare the IV estimator which exploits the refractory period with the CCH method given by \cref{eq:ptrans} which ignores network confounding. To indicate the amount of connections which are falsely attributed a non-zero weight we calculated the amount of false positives. This was given as the percentage of estimated synapses larger than $ 0.05 $ where the true weight was $ 0 $, finding $ 13.3\% $ for CCH and $ 0.2\% $ for the IV estimator; see \cref{fig:network-class}\labelcref{fig:network-class:1}. In addition we compared the size of the estimations at false positive instances and found that the CCH have slightly higher median. The IV approach, while not being perfect, thus outperforms the simple CCH approach.

It might be that modeling refractory periods in the context of a naive regression estimates connectivity equally well as the IV method. We thus performed a logistic regression as seen in \cref{fig:network-class}\labelcref{fig:network-class:1} denoted LOGIT. As seen here, LOGIT performs even worse than CCH showing that it really helps to use the refractory period as an instrumental variable. To further evaluate the methods we calculated false negatives as instances where the true weight is non-zero but estimated to be equal to zero in \cref{fig:network-class}\labelcref{fig:network-class:2} showing that the CCH and IV estimators performs equally well. Finally we wanted to evaluate the estimated weights as a function of true weights shown in \cref{fig:network-class}\labelcref{fig:network-class:3,fig:network-class:4} after 30000 trials. The IV estimator yields a good prediction $ r^2 = 0.55 $, while the CCH estimator performs surprisingly bad with $ r^2 = 0.002 $. Utilizing refractory periods as an (imperfect) instrumental variable considerably improves estimates.

\begin{figure}
\makeatletter
\renewcommand\p@subfigure{}
\makeatother
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{false_positive}
\caption{} \label{fig:network-class:1}
\end{subfigure}\hfill
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{false_negative}
\caption{} \label{fig:network-class:2}
\end{subfigure}
% * <tristan.stoeber@posteo.net> 2018-06-24T21:29:41.690Z:
% 
% Why is logit not considered for false negatives?
% 
% ^.

\medskip
\begin{subfigure}{0.46\textwidth} \includegraphics[scale=1]{fit_wald_pos}
\caption{} \label{fig:network-class:3}
\end{subfigure}\hfill
\begin{subfigure}{0.6\textwidth} \includegraphics[scale=1]{fit_cch_pos}
\caption{} \label{fig:network-class:4}
\end{subfigure}
\caption{{\bf False estimates and goodness of fit}. False positives are shown in \labelcref{fig:network-class:1} for the cross correlation histogram (CCH) method, logistic regression (LOGIT) and the IV estimator. False negatives for CCH and IV is shown in \labelcref{fig:network-class:2}. Positive estimates of weight as a function of true weight is scattered for the IV estimator in \labelcref{fig:network-class:3} and CCH in \labelcref{fig:network-class:4} color coded by the size of perturbation intensity. Shaded area show the 95 \% confidence interval calculated by bootstrapping.
\label{fig:network-class}}
\end{figure}

\section{Discussion}
%% maybe add somewhere that using other recorded neurons may be useful as well.

>>>>

\documentclass[11pt]{article}
\usepackage{amsmath,array}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{breakcites}
\usepackage{graphicx,subcaption}
\usepackage{todonotes} %[disable]
\usepackage{cleveref}
\usepackage{tikz}
\usepackage{placeins} %FloatBarrier
\usepackage{booktabs}
\usepackage{natbib}
\usetikzlibrary{arrows,shapes,calc}
\newcommand{\abs}[1]{\left|{#1}\right|}
\newcommand{\pde}[2]{{\frac{\partial {#1}}{\partial {#2}}}}
\newcommand{\pdde}[2]{\frac{\partial^2 {#1}}{\partial {#2}^2}}
\newcommand{\de}[2]{\frac{\mathrm{d} #1}{\mathrm{d} #2}}
\newcommand{\dde}[2]{\frac{\mathrm{d}^2 {#1}}{\mathrm{d} {#2}^2}}
\newcommand{\intdef}[4]{\int\limits_{#1}^{#2} #3\, \mathrm{d}{#4}}
\newcommand{\mean}[1]{\left<{#1}\right>}
\crefname{appendix}{Appendix}{appendices}
\crefname{section}{Section}{sections}
\crefname{table}{Table}{tables}
\Crefname{table}{Table}{Tables}
\crefname{equation}{Eq.}{eqs.}
\Crefname{equation}{Equation}{Equations}
\crefname{figure}{Fig.}{figs.}
\Crefname{figure}{Figure}{Figures}
\crefname{subfigure}{Fig.}{figs.}
\Crefname{subfigure}{Figure}{figures}
\captionsetup[subfigure]{labelformat=simple}
\renewcommand{\thesubfigure}{(\alph{subfigure})}
\renewcommand{\vec}[1]{\boldsymbol{#1}}
%opening
\title{Inferring causal connectivity from pairwise recordings and optogenetics}
%\title{The need to combine optogenetics with neural recordings to estimate causal connectivity}
\author{Mikkel Elle Lepper\o{}d, Tristan St$\ddot{\text{o}}$ber, Konrad Paul Kording}

@@

\begin{abstract}
\noindent Neuroscientists often want to identify how neurons causally interact. Perturbations with optogenetics are a popular way of quantifying such interactions. However, even focused light will typically stimulate multiple cells, which produces a confound - we cannot know which of the stimulated neurons affected the activity of a given target neuron. Here we show how this confound produces large biases in estimates of connectivity, and how we can reduce these biases by a data analysis strategy that exploits the absolute refractory period of neurons. We use the instrumental variable (IV) framework from economics which estimates causal strengths by exploiting a variable that affects only one of the variables, here the presynaptic neuron. The interaction between optogenetic stimulation and the absolute refractory period produces a weak almost random signal (IV) that we exploit to solve the causal inference problem. In a simple simulation of integrate and fire neurons, we find that instrumental variable techniques estimate connectivity better than naive techniques (99.8\% vs 0.6\% correct). Our instrumental variable approach is robust and easy to implement in the context of current experimental paradigms.
\end{abstract}

\section{Introduction}

\todo[inline]{Cite Konrad's new paper}
% neuroscience experiments are observational which implicates a certain set of statistical tools
% spike timing cannot give a causal estimate of connectivity, thus we need to perturbe the system
%a quest for causality and optogenetics is a main tool
% We want mechanism. Seriously. Neuroscience is to a large part experimentally driven.
%set up the drive to causality
We want to understand the mechanisms or causal chains that give rise to activity in the brain, to perception, action, and cognition. For such an understanding it is not sufficient to know the correlations between variables or even be able to predict them. After all, there can be many ways how the same activities come about by distinct causal chains \citep{drton2011global, peters2017elements}. Identifiability of causal networks is complicated when it comes to brain data. Complex systems such as the brain are hard to understand because of the numerous ways the contributing elements may interact internally \citep{Jonas2017}. Thus, brain data which is almost always observational, will almost never satisfy the criteria needed for identifiability \citep{pearl2009causality}. While observing correlations within the system is usually relatively easy, transitioning from observed correlations to a causal or mechanistic understanding is hard or, more commonly, impossible.  Getting at such an understanding in the human brain is nearly impossible as it contains approximately 86 billion neurons \citep{azevedo2009equal}, each of which influences many other neurons. Only under certain assumptions about nonlinearity or noise sources does a fully observed system become identifiable \citep{daniusis2012inferring,shimizu2006linear}. Even if we could record all neurons at the same time, estimating causality and producing a mechanistic understanding would be extremely challenging.

@@

%optogenetics is totally nonlocal
Interpreting the results of optogenetic stimulation in terms of causal interactions is difficult. The expression of opsins are often restricted to a population of neurons, but regular optogenetic stimulation will still affect many neurons simultaneously. In \textit{in vivo} settings it is generally impossible to direct photons to only one neuron. Hence, the stimulus will produce a distributed pattern of activity. For example, if we focus a stimulation beam on one neuron, there will be a cone of light in front of and behind the neuron, which comes from the rays of light from the lens focused on the cell and again coming out on the backside. This distributed pattern of stimulation produces activity which then percolates through the network of neurons. Thus any post-synaptic activity induced by stimulation could in principle come from any of the stimulated neurons introducing problematic confounders.

@@

\section{Results}
\subsection{Optogenetics is not local}
Optogenetics is generally seen as a perturbation method that by and large affects neurons in close proximity of the light source. However, it can be questioned if this is the correct way of conceptualizing the spatial effect of stimulation. The stimulation effects depend on multiple factors. Firstly, light intensity and opsin density are important as more light and ion channels will cause a stronger effect on each cell. Secondly, the number of potentially stimulated neurons is critical as more neurons will have a larger impact on the overall population activity. Lastly, physiological properties of the cells are important as light may e.g. only have a strong effect on spiking activity when the membrane potential of the cell is sufficiently close to the firing threshold. The induced effect of optogenetic stimulation as a function of distance should be given by a relation of the four parameters: light intensity, spatial distribution of neurons, distributions of membrane potential across neurons, and opsin distribution across neurons.

To estimate the light intensity we calculated the spatial extent of laser light delivered by fiber-optics under plausible experimental conditions according to \citet{Aravanis2007}; see \cref{sec:method:opto}. This modeling of light intensity yield an approximately $ 1/r^2 $ reduction with distance $ r $ from the stimulation site \cref{fig:concept} (cyan line). This is explained by the surface of the 3D shell growing with $4\pi r^2$ and photons will be roughly isotropic (but see \citep{thunemann2018does}) beyond the scattering length \cref{fig:concept} (inset). The same number of photons has to cross each of the spheres around the stimulation location unless they are absorbed. The density of photons thus decreases rapidly with distance.

@@

To estimate the effect of stimulation, the functionality underlying spiking activity needs to be considered. This can largely be characterized by the  distribution of membrane potentials across neurons. Surprisingly, this distribution has been observed to be symmetrically distributed and relatively flat \citep{pare1998impact,destexhe1999impact,rudolph2006use}. The response that is expected due to a pulse of light that induces a charge $Q$ should be proportional to the density of neurons whose membrane potential sit within a $Q/C$ range of the threshold ($C$ is the capacitance). The fact that the distribution of membrane potentials is relatively flat (the density close to the threshold is generally within an order of magnitude of the density of its mode) suggests that the spiking response to a perturbation for any neuron is roughly proportional to the induced photo current. The peak amplitude of the photo current relates non-linearly to the light intensity and can be modelled by the Hill equation \citep{wang2007high}. Moreover, assuming that opsins are evenly distributed along the neuron, the opsin density in the neuron population will be relatively even. Based on this, we calculate the overall stimulation effect to be the product of the number of neurons in a spherical slice and the peak amplitude photo current which actually increases with distance (up to the distance where absorption becomes important) \cref{fig:concept} (blue line). Thus, optogenetic stimulation utilizing single photon activation does not produce a localized effect.

\begin{figure}\includegraphics[scale=1]{opto-powerlaw}
\caption{{\bf Spatial extent of optogenetic stimulus}. Due to scattering and geometric loss the light intensity (I, cyan line) with an intensity of $10 mW/mm^2$ exiting the optogenetic fiber follows approximately an inverse square law $ r^{-2} $ where $ r $ is the distance from the fiber. If neurons are uniformly distributed the number of affected neurons increase by $ r^{2} $ (N, black line). The total photo current (sum of single neuron currents) thus increases by distance due to the nonlinear relation between photo intensity and photo current (P, blue line). \label{fig:concept}}
\end{figure}
\FloatBarrier
\subsection{Confounding as a problem for the estimation of causal effects}
When we stimulate many neurons at the same time, and observe a post-synaptic neuron to be active after our stimulation, it is hard to know which of the stimulated neurons produced the activity. To illustrate such confounding effects we simulated a network comprised of three neurons (A,B, and C) \cref{fig:intro}\labelcref{fig:intro:1}. The neurons receive Poisson spike trains and have added Gaussian white noise to the membrane potential. We also gave them interactions, where spikes of neuron B increase the probability of firing for neuron C but there are no other interactions.
Finally, we allowed simulated optogenetic stimulation to affect neurons A and B (but not C). We thus have a simple system for exploring questions of causality.

@@

\medskip
\begin{subfigure}{\textwidth}\centering \includegraphics[scale=1]{xcorr}
\caption{} \label{fig:intro:3}
\end{subfigure}
\caption{{\bf A,B,C network}. A sketch of the simple network containing three neurons %depicted in \labelcref{fig:intro:3} 
shows stimulation configuration with blue laser light and the connections with arrows (a). The neurons A and B are stimulated 1000 trials and the corresponding peristimulus time histogram are shown in \labelcref{fig:intro:2} upper panel with a raster plot in the lower panel. Cross correlation histograms (CCHs) are shown in \labelcref{fig:intro:3} where horizontal and vertical axes represents time lag in ms and counts of coincident spikes in bins of 1 ms. \label{fig:intro}}
\end{figure}
\FloatBarrier
\subsection{Instrumental variables to resolve confounding}
In order to estimate the actual influence of stimulation of a neuron on post-synaptic neurons we need something that can distinguish the influence of one stimulated neuron from the influence of another stimulated neuron. We would thus need something that affects the stimulation effect separately between neurons. Arguably, refractoriness is such a variable. If a neuron is in its absolute refractory period, then no amount of stimulation will make it spike. This gives us an interesting way of getting at causality, by comparing the network state between a time when a neuron is able to spike and a time where the neuron is unable to spike.

@@

We can now investigate if the use of an instrumental variable gives a better estimate of connectivity strength than simply analyzing the lagged correlations by means of the CCH; see \cref{eq:ptrans}. We thus use the IV estimator \cref{eq:wald} on the three neuron system (\cref{fig:cchvswald}\labelcref{fig:cchvswald:1,fig:cchvswald:2}). It converges to the correct causal conclusions that the weights $ w_{BC} = 0.2 $ and $ w_{AC} = 0 $. For such a simple system, it produces meaningful estimates of the causal interactions between neurons.
\tikzset{line/.style = {draw, -latex', ultra thick}}
\begin{figure}
\makeatletter
\renewcommand\p@subfigure{}
\makeatother
\begin{subfigure}{0.485\textwidth} \centering
\begin{tikzpicture}
    \node (u) {$ S $};
    \node [above = of u] (x) {$ A/B $};
    \node [above right = of u] (y) {$ C $};
    \node [left = of x] (sr) {$ A_r/B_r $};
    \path [line] (x) -- (y);
    \path [line] (u) -- (x);
    \path [line] (u) -- (y);
    \path [line] (sr) -- (x);
\end{tikzpicture}
\caption{} \label{fig:cchvswald:0}
\end{subfigure}\hfill
\begin{subfigure}{0.485\textwidth} \centering
\includegraphics[scale=1]{wald_legend}
\end{subfigure}\medskip

\begin{subfigure}{0.485\textwidth} 
\includegraphics[scale=1]{wald_BC}
\caption{} \label{fig:cchvswald:1}
\end{subfigure}\hfill
\begin{subfigure}{0.485\textwidth} 
\includegraphics[scale=1]{wald_AC}
\caption{} \label{fig:cchvswald:2}
\end{subfigure}
\caption{{\bf Instrumental variable estimation (IV) of connectivity}. \labelcref{fig:cchvswald:0} In an instrumental variable estimation procedure we use a variable that is assumed to be random (here refractoriness) which influences a variable of interest (here spiking) and to use this influence to infer the causal interaction of that variable on other variables (here spiking of A or B onto C). \labelcref{fig:cchvswald:1} A popular estimation approach for IVs,  the Wald technique correctly estimates causal connectivity in the A,B,C neural network using the refractory period. Subfigure \labelcref{fig:cchvswald:0} shows a path diagram \citep{wright1921correlation} between the upstream neuron $ A $ or $ B $, the stimulation $ S $, the post-synaptic neuron $ C $ and the IV as $ A_{r} $ or $ B_{r} $. Arrows represent associations where $ S $ is associated with $ A,B $ and potentially also with $ C $ both directly and through $ A,B $. The IV estimator calculated by \cref{eq:wald} converges to $ \hat{\beta}_{BC} \approx 0.2, \hat{\beta}_{AC} \approx 0 $ after approximately $ 5000 $ trials as seen in \labelcref{fig:cchvswald:1,fig:cchvswald:2} respectively. 
 \label{fig:cchvswald}}
\end{figure}
\FloatBarrier
\subsection{Larger simulated networks}
The interacting neurons in a biological network exhibit inhibition and interact in many kinds of ways. To evaluate the IV method in a more meaningful setting we thus simulated a recurrent neural network consisting of 1250 randomly connected leaky integrate and fire neurons where 250 had inhibitory synapses. The network was tuned to be in an asynchronous regime; see \cref{fig:state}\labelcref{fig:CC} with log-normally distributed synaptic weights according to patch clamp experiments \citep{Sayer1990,Mason1991}; see \cref{fig:state}\labelcref{fig:syn_dist} and \cref{tab:params} for parameters. Further, we selected 800 excitatory neurons for stimulation and gave each neuron a random spatial distance from the simulated optogenetic stimulus. The stimulus intensity was then set according to \cref{eq:intensity} with a maximum of 10 pA, and was constant throughout trials. The trial onset had a temporal Poisson distribution with period 100 ms and was further clipped between 100-150 ms. For weight estimates we randomly selected among the excitatory population 100 stimulated neurons and 100 unstimulated neurons.

@@

\begin{subfigure}{\textwidth} 
\includegraphics[scale=1]{mse}
\caption{} \label{fig:error:mse}
\end{subfigure}
\caption{{\bf Mean square error (MSE) of IV estimator in a large network}. \labelcref{fig:error:raster} Excitatory neurons (in blue raster) are stimulated with varying intensity indicated by the y-axis in lower panel, upper panel shows inhibitory neurons (red raster) where y-axis indicate random neuron-id. \labelcref{fig:error:mse} The IV- and CCH estimator is evaluated for the asynchronous recurrent neural network at three different amounts of relative inhibition $ g $. The MSE as a function of number of trials is shown on a logarithmic scale where the slopes was found to be $ -0.67, -0.69, -0.45 $ for $g = 9.9, 4.4, 3.0$ respectively. \label{fig:error}}
\end{figure}

We want to compare the IV estimator which exploits the refractory period with the CCH method given by \cref{eq:ptrans} which ignores network confounding. We defined hit rate as the relative amount of stimulations in non-refractory states and required a minimum of 90$\%$. To indicate the amount of connections which are falsely attributed a non-zero weight we calculated the amount of false positives. This was given as the percentage of estimated synapses larger than $ 0.05 $ where the true weight was $ 0 $, finding $ 99.4\% $ for CCH and $ 0.2\% $ for the IV estimator; see \cref{fig:network-class}\labelcref{fig:network-class:1}. In addition we compared the size of the estimations at false positive instances and found that the CCH have significantly higher median than IV (p=0, difference ($\Delta$) = 0.17, permutation resampling \citep{wassermann2006all}). The IV approach, while not being perfect, thus outperforms the simple CCH approach.

It might be that modeling refractory periods in the context of a naive regression estimates connectivity equally well as the IV method. We thus performed a logistic regression as seen in \cref{fig:network-class}\labelcref{fig:network-class:1} denoted LOGIT. As seen here, LOGIT performs even worse than CCH showing that it really helps to use the refractory period as an instrumental variable (p=0, $\Delta$ = 0.7, permutation resampling). To further evaluate the methods we calculated false negatives as instances where the true weight is non-zero but estimated to be equal to zero in \cref{fig:network-class}\labelcref{fig:network-class:2} showing that the CCH and IV estimators performs equally well (p=0.78, $\Delta$ = 0.021, permutation resampling). Finally we wanted to evaluate the estimated weights as a function of true weights shown in \cref{fig:network-class}\labelcref{fig:network-class:3,fig:network-class:4} after 30000 trials. The IV estimator yields a good prediction ($ R^2 = 0.845 $), while the CCH mainly estimates the strength of stimulation while true weights are poorly estimated ($ R^2 = 0.048 $). Utilizing refractory periods as an (imperfect) instrumental variable considerably improves estimates.

\begin{figure}
\makeatletter
\renewcommand\p@subfigure{}
\makeatother
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{false_positive}
\caption{} \label{fig:network-class:1}
\end{subfigure}\hfill
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{false_negative}
\caption{} \label{fig:network-class:2}
\end{subfigure}
